Efficient Computation of Extended Sources

### 2021-06-08 repro/papers/sseprecond/paper.tex
interim

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. The first row of
\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
repeat for the reader's convenience:
\begin{equation}
  \label{eqn:lamidea}
  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}

@@

Figure \ref{fig:preddinvhshh0} shows the pressure source gather produced
from the pressure data gather in Figure \ref{drecphh0} by application
of the operator on the right-hand side of
formula \ref{eqn:lamident}. The scale is the same as for Figures
\ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact the former
appears nearly identical, and both appear to be dip-filtered versions
of the latter. The difference plot (Figure \ref{fig:ddiffinvhshh0},
using the same color scale, shows that the similarity is quantitative.

Figures \ref{fig:preddinvhsll0} and \ref{fig:ddiffinvhsll0} accomplish
the same comparison for the $\Lambda^+_{z_s}$ approximation in
equation \ref{eqn:lamident}. A deeper study of the pressure-to-source
operator shows that it is approximately dependent only on the model
coefficients near the source surface ($z=z_s$ in this case). Since the
homogeneous and lens models are identical near this surface, it is
unsurprising that these figures are very close to the previous two.

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamident} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Since successful preconditioning requires only approximate inversion,
use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
still yield a working preconditioner, and the former can be transposed
to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_r,z_s})^T\Pi_0^T V^{+}_{z_s,z_r}
 \end{equation}

>>>>

Computations of $\Lambda^{\pm}_{z_s}$ and its transpose are clearly critical steps in an
implementation of the PCG algorithm outlined in the preceding section.
Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ constructed above. The first row of
\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
repeat for the reader's convenience:
\begin{equation}
  \label{eqn:lamidea}
  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}

@@

Figure \ref{fig:preddinvhshh0} shows the pressure source gather
produced from the pressure data gather in Figure \ref{dsrcphh0} by
application of the operator on the right-hand side of formula
\ref{eqn:lamident} to the source surface pressure gather depicted in
\ref{fig:dsrcphh0}. The homogeneous model is used in all propagations
implicit in the prescription \ref{eqn:lamident}. The scale is the same
as for Figures \ref{fig:dinvhshh0} and \ref{fig:dhshh0}. In fact
Figure \ref{fig:preddinvhshh0} appears to be nearly identical to
Figure \ref{fig:dinvhshh0}, and both appear to be dip-filtered
versions of Figure \ref{fig:dhshh0. The difference plot (Figure
  \ref{fig:ddiffinvhshh0}, using the same color scale, shows that the
  similarity is quantitative. The recovered source gather (Figure
  \ref{fig:preddinvhshh0}) also generates the same acoustic fields as
  the point source at $z=z_d$: the pressure gather at $z=z_r=1000$ m
  depth is shown in Figure \ref{fig:dpredhsrecphh0}, and its
  difference with the point source gather (Figure \ref{fig:drecphh0})
  plotted on the same color scale in Figure
  \ref{fig:ddiffpredhsrecphh0}.

Figures \ref{fig:preddinvhsll0}, \ref{fig:ddiffinvhsll0},
\ref{fig:dpredhsrecpll0}, and \ref{fig:ddiffpredhsrecpll0} accomplish
the same comparison for the $\Lambda^+_{z_s}$ approximation in
equation \ref{eqn:lamident}, with all propagation taking place in the
lens model.

A deeper study of the pressure-to-source
operator (Appendix B) shows that it is approximately dependent only on the model
coefficients near the source surface ($z=z_s$ in this case). Since the
homogeneous and lens models are identical near this surface, it is
unsurprising that these figures are very close to the previous two.
However an even more useful observation is that the calculations in
the approximation \ref{eqn:lamident} could just as well be carried out
in a much smaller region around the source surface, and produce a
result that is functionally identical in that it will serve as a
source for the same acoustic fields globally, with small error. In
effect, equation \ref{eqn:lamident} involving propagation from source
($z=z_s$) to receiver ($z=z_r$) surfaces is altered by replacing $z_r$
with a receiver datum $z_s+\Delta z$ considerably closers to $z_s$:
\begin{equation}
  \label{eqn:lamnear}
  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_z_s+\Delta z}\Pi_1^T.
\end{equation}
Using a receiver datum closer to the source surface has two favorable consequences:
\begin{itemize}
\item The computational domain can be smaller than is necessary to
  simulate the target data, as it need only contain the source
  surface and the receiver datum implicit in
  equation \ref{eqn:lamident}. This shrinkage of the computational
  domain can lead to substantial improvements in computational
  efficiency.
\item Since the receiver data may be chosen much closer to the
  source surface that is the case for the target data, the effective
  aperture active in the relation \ref{eqn:lamident} can be much
  larger, producing an estimated source gather much less affected by
  aperture limitation.
\end{itemize}

To illustrate these points, I have re-computed the source gather
estimates shown in Figures \ref{fig:preddinvhshh0} and
\ref{fig:preddinvhsll0}. In place of the receiver geometry in the
target data (Figures \ref{fig:drecphh0} and \ref{fig:drecpll0}, I
place the receiver array at $z_s+\Delta z=2900$ m depth, just 100 m above the
source surface at $z_s=3000$ m. For the discretization used to create
the examples shown so far, that is just a 5 gridpoint difference in
depth, as opposed to 100 gridpoints between the source and receiver
depths for examples such as shown in Figure \ref{fig:recphh0} or
\ref{fig:drecphh0}. The resulting approximation to $\Lambda_{z_s}^+$
via equation \ref{eqn:lamident} for this configuration is evaluated in
the same way as before in Figures \ref{fig:preddnshshh0}, \ref{fig:ddiffnshshh0},
\ref{fig:dprednshsrecphh0}, and \ref{fig:ddiffprednshsrecphh0}
(homogeneous model) and Figures \ref{fig:preddnshsll0}, \ref{fig:ddiffnshsll0},
\ref{fig:dprednshsrecpll0}, and \ref{fig:ddiffprednshsrecpll0} (lens
model). The effect of aperture limitation is clearly diminished: the
second figure in each series compares the full-aperture pressure source
gathers (Figures \ref{fig:dhshh0.su} and \ref{fig:dhsll0.su} with the
approximated image of the corresponding pressure gather (Figure
\ref{fig:dsrcphh0}) under the approximation to $\Lambda_{z_s}^+$, and
the last figures show that the approximated source gathers accurately
predict the point-source pressure gather at the receiver datum
$z_r=1000$ m.

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = -8(V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamnear} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Since successful preconditioning requires only approximate inversion,
use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
still yield a working preconditioner, and the former can be transposed
to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   (\tilde{\Lambda}^+_{z_s})^T = -8 \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
     z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}
 \end{equation}
Thus the model space weight operator introduced in the last section is
replaced by
\[
  \tilde{W}_m = \frac{1}{2}(\tilde{\Lambda}^+_{z_s} +
  (\tilde{\Lambda}^+_{z_s})^T)
\]
\[
  = -8 \left((V^{+}_{z_s,z_s+\Delta z})^T\Pi_0 {\cal
    S}^+_{z_s,z_s+\Delta z}\Pi_1^T \right.+
\]
\[
  \left. \Pi_1 ({\cal S}^+_{z_s,z_s+\Delta
      z})^T\Pi_0^T V^{+}_{z_s,z_s+\Delta z}\right)
\]
\begin{equation}
  \label{eqn:wcomp}
  = -4(\Pi_1 ({\cal S}^+_{z_s,z_s+\Delta z})^T(\Pi_0^T\Pi_1 + \Pi_1^T\Pi_0) {\cal S}^+_{z_s,z_s+\Delta z}\Pi_1^T)
\end{equation}
